\addchap{Acknowledgements}

First and foremost I would like to thank my supervisor John Wickerson for his
continued guidance and patience throughout the PhD.  He has allowed me to
explore freely, and was generous with his time whenever I needed help or advice.
He has also always encouraged me to present my work in diverse venues, leading
to many interesting discussions and connections.  His wonderful ability to
understand technical descriptions and expose the core contribution clearly and
creatively has been invaluable, and has taught me how to communicate my ideas
clearly.  I would also like to thank my examiners George Constantinides and Xavier Leroy for the many thought-provoking discussions and helpful suggestions that lead to a much clearer dissertation and many new ideas to explore.

Next, I would like to thank my colleagues in the Circuits and Systems group for
many interesting discussions and creating an enjoyable atmosphere in the lab:
Jianyi for a joint journey through the PhD, Nadesh for introducing me to
high-level synthesis and helping me run them in the first place, Aditya and
Diederik for many discussions about compilers over drinks, my coauthors James
Pollard, Zewei and Michalis for all their help on our publications, and Kate,
Roy, Divyansh and Mariano for being amazing flatmates.  Finally, I would like to
thank Alex Dalton, Alex Montgomerie-Corcoran, Ben Biggs, Ben Chua, Cheng, Dan, Erwei, Guoxuan, Ian, Marta, Quentin, Sina and Zhewen
for many insightful discussions and advice, as well as help working with FPGAs
and hardware tools.

I also gratefully acknowledge the Cyber Security Centre (NCSC) for funding this
PhD through the Research Institute on Verified Trustworthy Software Systems
(VeTSS).

I have also had the great pleasure to be able to work in the Celtique group at
Irisa in Rennes as a visiting researcher over the summer in 2022.  I would like
to thank Sandrine Blazy and Delphine Demange for their deep insights into the
CompCert compiler and for countless interesting discussions on different
verification techniques and how these could be integrated into CompCert.

I would also like to sincerely thank Sandro Stucki and Bor-Yuh Evan Chang for
their invaluable guidance during my internship at Amazon Prime Video, as well as
my colleagues during my time there with whom I had enlightening discussions
with: Daniel, Francesco, Franco, Horia, Ilina, Ioannis, Pauline, Philipp, Sarek,
Stefan, and Vlad.  Additionally, I would like to thank Stefan Zetzsche for all
his help understanding Dafny.
%with: Daniel Schoepe, Francesco Parolini, Franco Raimondi, Horia Druliac, Ilina
%Stoilkovska, Ioannis Boutsis, Pauline Bolignano, Philipp Renz, Sarek Høverstad
%Skotåm, Stefan Giba, Vlad Nicolicin Georgescu.  Additionally, I would like to
%thank Stefan Zetzsche for all his help while working with Dafny.

I would also like to thank my family and friends for their invaluable support
and understanding during the PhD.  Thank you to my father and especially my mother for their continued
support, helping me get through stressful times.

Finally, I would like to thank my partner and best friend Nikita for her
continued support.

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "thesis"
%%% TeX-engine: luatex
%%% End:
